---
title: "A1 Soln"
author: "Xiangyu Kong"
date: "15/01/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 60)
)
```


# Question 1


## Read Data

```{r 1_data}
# Read the crime show ratings data
crime_show_file = "crime_show_ratings.RDS"
crime_show_data = readRDS(crime_show_file)
```

## Question 1.a

Let $y_i$ denote season rating for sample $i$.

Let $x_{i, 2000}$ be indicator variable that is set to 1 if the decade for the sample $i$ is 2000, 0 otherwise.

Let $x_{i, 2010}$ be indicator variable that is set to 1 if the decade for the sample $i$ is 2010, 0 otherwise.

Equation for linear model:

$$
y_{i} = \beta_0 + \beta_1 x_{i, 2000} + \beta_2 x_{i, 2010} + \epsilon_{i}
$$

Anova Assumptions:

1. Errors ($\epsilon_i$) are independent
2. Errors are normally distributed with $E[\epsilon_i] = 0$
3. Errors have constant variance $var[\epsilon] = \sigma^2$


## Question 1.b

The hypotheses for ANOVA are listed and can be described as follows:

- $H_0$: $\mu_{1990} = \mu_{2000} = \mu_{2010}$: 
The mean season rating for crime shows are the same across different decades. I.e. Different decades does not have effect over season ratings
- $H_1$: at least one mean is different from the others:
The mean season rating for crime shows are different across different decades. I.e. Different decades has at least some effect over season ratings


\newpage
## Question 1.c

```{r 1_c_boxplot}
# Side by side box plots
crime_show_data %>%
  ggplot(aes(x = decade, y = season_rating)) +
  geom_boxplot() +
  ggtitle("Boxplots of average rating by decade for crime TV shows")
```

\newpage
```{r 1_c_histogram}
# Facetted histograms
crime_show_data %>%
  ggplot(aes(x = season_rating)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ decade) +
  ggtitle("Histograms of average rating by decade for crime TV shows")
```

The box plot provides a better visualization of the data because it shows comparison across three decades' basic statistics (maximum, minimum, quartiles, median) side by side. 

On the other side, with the histograms, it is harder to tell which decade has a higher rating because it only provides visualization over frequencies within each decade, and provides a relatively poor visualization for comparing between different decades.

One improvement for the box plot could be to cleaning the data before plotting. In the plot, we observe that there are some outliers, especially with 1990 and 2010. Removing those outliers may provide a even better visualization.

According to the box plot, we can see that the boxes are roughly on the same level. There is no sign of extremely skewed data except for some outliers, so their means are similar to the median (all around 8). Thus it does not suggest a significant difference between the means.


\newpage
## Question 1.d

``` {r 1_d_one_way_anova}
one_way_anova <- aov(season_rating ~ decade, data = crime_show_data)

summary(one_way_anova)
```

From the one way anova we can see that the p-value for the F-test is 0.238. We can interpret this as: The probability of observing the current sample given the assumption that the three decades having the same mean season ratings  is 0.238.


## Question 1.e

``` {r 1_e_sd_plot}
plot(one_way_anova, 1)
```

\newpage
``` {r 1_e_plot_2}
plot(one_way_anova, 2)

crime_show_data %>%
  group_by(decade) %>%
  summarise(var_rating = sd(season_rating) ^ 2)
```

The first plot is the Residual vs Fitted plot. The plot shows that except for some ouliers, the residuals are roughly randomly scattered around the 0-line, and does not indicate any pattern. This shows that the data follows a linear relationship, have equal error variances, and have a few outliers.

The second plot is the normal q-q plot. From the plot, we can see that except for points 19, 165, and 43, the points form a relatively straight line, indicating that the data follows a normal distribution with a few outliers.

From the standard deviations, we calculate that the ratio of the largest within-group and biggest within-group variance estimate is $\dfrac{0.480}{0.203} = 2.365 < 3$. According to the rule of thumb from Dean and Voss, the assumption for equality of variances is satisfied.


\newpage
## Question 1.f

``` {r 1_f_linear_model}
lm1 = lm(season_rating ~ decade, data = crime_show_data)

summary(lm1)
```

The linear model can be expressed as 

$$
y = \beta_0 + \beta_1 x_{2000} + \beta_2 x_{2010}
$$

where $y$ is the season rating, $\beta_i$s are the coefficients, $x_{2000}$ is the indicator variable for decade 2000, and $x_{2010}$ is the indicator variable for decade 2010.

$\beta_0$ is the intercept of the regression line, which is equal to the sample mean for decade 1990.

$\beta_1$ is the amount of score increase when the indicator variable $x_{2000}$ is set to 1.

$\beta_2$ is the amount of score increase when the indicator variable $x_{2010}$ is set to 1.

Then the sample mean for decade 1990 $\hat{\mu}_{1990} = \beta_0 = 7.9222$.

The sample mean for decade 2000 $\hat{\mu}_{2000} = \beta_0 + \beta_1 = 7.9222 + 0.1368 = 8.059$.

The sample mean for decade 2000 $\hat{\mu}_{2010} = \beta_0 + \beta_2 = 7.9222 + 0.1938 = 8.116$.


\newpage
# Question 2


## Read Data

```{r 2_data}
# Read the crime show ratings data
smokeFile = 'smokeDownload.RData'
if (!file.exists(smokeFile)) {
  download.file('http://pbrown.ca/teaching/303/data/smoke.RData',
                smokeFile)
}
(load(smokeFile))

smokeFormats[smokeFormats[, 'colName'] == 'chewing_tobacco_snuff_or',
             c('colName', 'label')]

# get rid of 9, 10 year olds and missing age and race
smokeSub = smoke[which(smoke$Age > 10 & !is.na(smoke$Race)), ]
smokeSub$ageC = smokeSub$Age - 16
smokeModel = glm(
  chewing_tobacco_snuff_or ~ ageC + RuralUrban + Race + Sex,
  data = smokeSub,
  family = binomial(link = 'logit')
)
knitr::kable(summary(smokeModel)$coef, digits = 3)

logOddsMat = cbind(est = smokeModel$coef, confint(smokeModel, level = 0.99))
oddsMat = exp(logOddsMat)
oddsMat[1, ] = oddsMat[1, ] / (1 + oddsMat[1, ])
rownames(oddsMat)[1] = 'Baseline prob'
knitr::kable(oddsMat, digits = 3)
```


## Question 2.a

The statistical model that corresponds to `smokeModel` is 

$$
Y_{i} \sim Binomial(N_{i}, \mu_{i})
$$
$$
h(\mu_i) = log(\frac{\mu_i}{1 - \mu_{i}}) = X^{T}_i \beta
$$

where

- $\mu_{i}$ is the sample mean $E(Y_i)$
- $N_{i}$ is the population size 
- $h(\mu_{i})$ is the logit link function

Within $X_{i}$, there is an intercept of all 1's, a numeric variable of age centered around 16, and indicator variables for Region (Rural / Urban), Race (Black, Hispanic, Asian, Native, Pacific), and Sex (Male, Female).


## Question 2.b

For the baseline prob row,

- The value under `est` refers to the probability of observing the subset of individuals who are 16-year-old urban white males that have used chewing tobacco, snuff, or dip on 1 or more days in the past 30 days.
- The value under $0.5\%$ and $99.5\%$ represents the $99\%$ confidence interval for the estimated probability


## Question 2.c

``` {r 2_c_}
newData = data.frame(
  Sex = rep(c('M', 'F'), c(3, 2)),
  Race = c('white', 'white', 'hispanic', 'black', 'asian'),
  ageC = 0,
  RuralUrban = rep(c('Rural', 'Urban'), c(1, 4))
)

smokePred = as.data.frame(predict(smokeModel, newData,
                                  se.fit = TRUE, type = 'link'))[, 1:2]

smokePred$lower = smokePred$fit - 3 * smokePred$se.fit
smokePred$upper = smokePred$fit + 3 * smokePred$se.fit
smokePred

expSmokePred = exp(smokePred[, c('fit', 'lower', 'upper')])
knitr::kable(cbind(newData[, -3], 1000 * expSmokePred / (1 + expSmokePred)),
             digits = 1)
```

The claim that rural white males are the group most likely to use
chewing tobacco is likely to be true. This is because the confidence interval of rural white males do not overlap with any other group, and have the highest fit value.

The claim that less than half of one percent of ethnic-minority urban
women and girls chew tobacco can neither be confirmed or denied base on the data. Considering black and asian urban female, the fitted probability of them chewing tobacco is in $99\%$ confidence interval of $(0.13\%, 0.42\%)$ and $(0.08\%, 0.68\%)$. We can see that the $99\%$ confidence interval for black urban women is less than half of one percent. However, the $99\%$ confidence interval for asian urban women includes $0.5\%$ of its population. Thus we cannot conclude the claim is true.


# Question 3


## Read Data

``` {r}
fijiFile = 'fijiDownload.RData'
if (!file.exists(fijiFile)) {
  download.file('http://pbrown.ca/teaching/303/data/fiji.RData',
                fijiFile)
}
(load(fijiFile))
```

``` {r}
# get rid of newly married women and those with missing literacy status
fijiSub = fiji[fiji$monthsSinceM > 0 & !is.na(fiji$literacy), ]
fijiSub$logYears = log(fijiSub$monthsSinceM / 12)
fijiSub$ageMarried = relevel(fijiSub$ageMarried, '15to18')
fijiSub$urban = relevel(fijiSub$residence, 'rural')
fijiRes = glm(
  children ~ offset(logYears) + ageMarried + ethnicity + literacy + urban,
  family = poisson(link = log),
  data = fijiSub
)
logRateMat = cbind(est = fijiRes$coef, confint(fijiRes, level = 0.99))
knitr::kable(cbind(summary(fijiRes)$coef,
                   exp(logRateMat)),
             digits = 3)
```

``` {r}
fijiSub$marriedEarly = fijiSub$ageMarried == '0to15'
fijiRes2 = glm(
  children ~ offset(logYears) + marriedEarly + ethnicity + urban,
  family = poisson(link = log),
  data = fijiSub
)
logRateMat2 = cbind(est = fijiRes2$coef, confint(fijiRes2, level = 0.99))
knitr::kable(cbind(summary(fijiRes2)$coef,
                   exp(logRateMat2)),
             digits = 3)
```

``` {r}
lmtest::lrtest(fijiRes2, fijiRes)
```


## Question 3.a

The statistical model that corresponds to `fijiRes` is 

$$
Y_{i} \sim Poisson(\lambda)
$$
$$
h(\mu_{i}) = h(\lambda_{i}) = log(\frac{\lambda_{i}}{O_{i}}) = X^{T}_i \beta
$$

where

- $\mu_{i}$ is the sample mean $E(Y_i)$
- $O_{i}$ is the offset term, which is the number of months since married.
- $h(\mu_{i})$ is the log link function

Within $X_{i}$, there is an intercept of all 1's, and indicator variables for Age range of marriage (15-18, 0-15, 18-20, 0-22, 22-25, 25-30, 30+), ethnicity (Fijian, Indian, European, Part Eruopean, Pacific Islander, Routman, Chinese, Others) + literacy (Yes, No) + urban (Rural, Suva, Other Urban).

The intercept represents predicted rates of children per month for the subset of females who are between 15 to 18 years old, Fijian, rural and literate.


## Question 3.b

The model `fijiRes2` is nested within `fijiRes`.

<!-- TODO: what does it mean by constraints on the vector of regression coefficients \beta in the restricted model? -->
`fijiRes2` can be viewed as `fijiRes` stripping ethnicity, and only retaining one age group indicator variable. The marriedEarly variable can simply be seen as the indicator variable for Age range of marriage 0-15. Thus the coefficients $\beta$ for `fijiRes2` will have 6 less rows and is a subset of the coefficients for `fijiRes`.


## Question 3.c

From the lmtest result above, we can see that the model being compared is `fijiRes2` and `fijiRes`.

`fijiRes` takes into account for women of different ages and their literacy.

`fijiRes2` is nested within `fijiRes`, and does not account for women of different ages above 15 years old, and their literacy.

The p-value for the test is $0.3834$. This is not significant enough to reject the null hypothesis that adding literacy and age range improves how well the model explains the data.

Thus the second claim is likely to be true.


<!-- \newpage -->
<!-- # Appendix -->

<!-- ```{r ref.label=knitr::all_labels(), echo = T, eval = F} -->
<!-- ``` -->
