---
title: "A1 Soln"
author: "Xiangyu Kong"
date: "15/01/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  tidy.opts = list(width.cutoff = 60)
)
```


# Question 1


## Read Data

```{r 1_data}
# Read the crime show ratings data
crime_show_file = "crime_show_ratings.RDS"
crime_show_data = readRDS(crime_show_file)
```

## Question 1.a

Let $y_i$ denote season rating for sample $i$.

Let $x_{i, 2000}$ be indicator variable that is set to 1 if the decade for the sample $i$ is 2000, 0 otherwise.

Let $x_{i, 2010}$ be indicator variable that is set to 1 if the decade for the sample $i$ is 2010, 0 otherwise.


Equation for linear model:

$$
y_{i} = \beta_0 + \beta_1 x_{i, 2000} + \beta_2 x_{i, 2010} + \epsilon_{i}
$$

Anova Assumptions:

1. Errors ($\epsilon_i$) are independent
2. Errors are Normally distributed with $E[\epsilon_i] = 0$
3. Errors have constant variance $var[\epsilon] = \sigma^2$


## Question 1.b

The hypotheses for ANOVA are listed and can be described as follows:

- $H_0$: The mean season rating for crime shows are the same across different decades. I.e. Different decades do not have effect on average season ratings
- $H_1$: Across different decades, least one mean is different from the others: The mean season rating for crime shows are different across different decades. I.e. Different decades have at least some effect on average season ratings


\newpage
## Question 1.c

```{r 1_c_boxplot}
# Side by side box plots
crime_show_data %>%
  ggplot(aes(x = decade, y = season_rating)) +
  geom_boxplot() +
  ggtitle("Boxplots of average rating by decade for crime TV shows")
```

\newpage
```{r 1_c_histogram}
# Facetted histograms
crime_show_data %>%
  ggplot(aes(x = season_rating)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ decade) +
  ggtitle("Histograms of average rating by decade for crime TV shows")
```

The box plot provides a better visualization of the data because it shows comparison across three decades' basic statistics (maximum, minimum, quartiles, median) side by side. 

On the other side, with the histograms, it is harder to tell which decade has a higher average rating because it only provides visualization over frequencies within each decade, and provides a relatively poor visualization for comparing between different decades.

One improvement for the box plot could be to sanitizing the data before plotting. In the plot, we observe that there are some outliers, especially with decades 1990 and 2010. Removing those outliers may provide a better visualization.

Another improvement could be adding a geom dot of the mean for each decade to the box plot as in week4's lab. This is because boxplot only shows quartiles instead of means. Adding the mean to the plot gives a more precise observation.

According to the box plot, we can see that the boxes are roughly on the same level. There is no sign of extremely skewed data except for some outliers, so their means are similar to the median (all around 8). Thus it does not suggest a significant difference between the means.


\newpage
## Question 1.d

``` {r 1_d_one_way_anova}
# one way anova
one_way_anova <- aov(season_rating ~ decade, data = crime_show_data)

summary(one_way_anova)
```

From the one way anova we can see that the p-value for the F-test is 0.238. We can interpret this as: The probability of observing the current sample given the assumption that the three decades having the same mean season ratings is 0.238. Using an $\alpha$ value of $5\%$, the p-value is not significant enough for us to reject $H_0$. Thus we cannot reject the statement that different decades do not have any effect on mean season ratings.


## Question 1.e

``` {r 1_e_plot_1}
# first plot of one way anova
plot(one_way_anova, 1)
```

\newpage
``` {r 1_e_plot_2}
# second plot for one way anova
plot(one_way_anova, 2)

# variance for different decades
crime_show_data %>%
  group_by(decade) %>%
  summarise(var_rating = sd(season_rating) ^ 2)
```

The first plot is the Residual vs Fitted plot. The plot shows that except for some ouliers, the residuals are roughly randomly scattered around the 0-line, and they do not indicate any pattern. This shows that the data follows a linear relationship, have equal error variances, and have a few outliers.

The second plot is the normal q-q plot. From the plot, we can see that except for points 19, 165, and 43, the points form a relatively straight line, indicating that the data follows a normal distribution with a few outliers.

From the standard deviations, we calculate that the ratio of the largest within-group and biggest within-group variance estimate is $\dfrac{s^{2}_{max}}{s^{2}_{min}} = \dfrac{s^{2}_{1990}}{s^{2}_{2000}} = \dfrac{0.480}{0.203} = 2.365 < 3$. According to the rule of thumb from Dean and Voss, the assumption for equality of variances is satisfied.


\newpage
## Question 1.f

``` {r 1_f_linear_model}
# linear model
lm_rating_decade = lm(season_rating ~ decade, data = crime_show_data)

summary(lm_rating_decade)
```


The linear model can be expressed as 

$$
y = \beta_0 + \beta_1 x_{2000} + \beta_2 x_{2010}
$$

where $y$ is the season rating, $\beta_i$s are the coefficients, $x_{2000}$ is the indicator variable for decade 2000, and $x_{2010}$ is the indicator variable for decade 2010.

$\beta_0$ (`intercept`) is the intercept of the regression line, which is equal to the sample mean for decade 1990.

$\beta_1$ (`decade2000`) is the amount of score increase when the indicator variable $x_{2000}$ is set to 1. I.e. the sample mean for decade 2000 will be $0.1368$ larger than that of decade 1990.

$\beta_2$ (`decade2010`) is the amount of score increase when the indicator variable $x_{2010}$ is set to 1. I.e. the sample mean for decade 2010 will be $0.1368$ larger than that of decade 1990.

Then the sample mean for decade 1990 $\hat{\mu}_{1990} = \beta_0 = 7.9222$.

The sample mean for decade 2000 $\hat{\mu}_{2000} = \beta_0 + \beta_1 = 7.9222 + 0.1368 = 8.059$.

The sample mean for decade 2000 $\hat{\mu}_{2010} = \beta_0 + \beta_2 = 7.9222 + 0.1938 = 8.116$.


\newpage
# Question 2


## Read Data

```{r 2_data}
# Read the crime show ratings data
smokeFile = 'smokeDownload.RData'
if (!file.exists(smokeFile)) {
  download.file('http://pbrown.ca/teaching/303/data/smoke.RData',
                smokeFile)
}
(load(smokeFile))

smokeFormats[smokeFormats[, 'colName'] == 'chewing_tobacco_snuff_or',
             c('colName', 'label')]
```


## Question 2.a

``` {r 2_a_binomial_glm}
# remove na values and unreasonable ages
smokeSub = smoke[which(smoke$Age > 10 & !is.na(smoke$Race)), ]
# center age around 16 years old
smokeSub$ageC = smokeSub$Age - 16

# binomial glm model
smokeModel = glm(
  chewing_tobacco_snuff_or ~ ageC + RuralUrban + Race + Sex,
  data = smokeSub,
  family = binomial(link = 'logit')
)
knitr::kable(summary(smokeModel)$coef, digits = 3)
```

\newpage
The statistical model that corresponds to `smokeModel` is 

$$
Y_{i} \sim Binomial(N_{i}, \mu_{i})
$$
$$
h(\mu_i) = log(\frac{\mu_i}{1 - \mu_{i}}) = X^{T}_i \beta
$$

where

- $Y_{i}$ is the number of people who has used chewing tobacco, snuff, or dip on 1 or more days in the past 30 days.
- $\mu_{i}$ is the probability of a single person used chewing tobacco, snuff, or dip on 1 or more days in the past 30 days.
- $N_{i}$ is the sample size for the people following the indicator variable assignment.
- $h(\mu_{i})$ is the logit link function.

Within $X_{i}$, there is an intercept of all 1's, a numeric variable of age centered around 16, and indicator variables for Region (Rural / Urban), Race (Black, Hispanic, Asian, Native, Pacific), and Sex (Male, Female).


## Question 2.b
``` {r 2_b_odds_table}
# odds table with 99% CI
logOddsMat = cbind(est = smokeModel$coef, confint(smokeModel, level = 0.99))
oddsMat = exp(logOddsMat)
oddsMat[1, ] = oddsMat[1, ] / (1 + oddsMat[1, ])
rownames(oddsMat)[1] = 'Baseline prob'
knitr::kable(oddsMat, digits = 3)
```

For the `Baseline prob` row,

- The value under `est` refers to the probability of observing the subset of individuals who are 16-year-old urban white males that have used chewing tobacco, snuff, or dip on 1 or more days in the past 30 days.
- The value under $0.5\%$ and $99.5\%$ represents the $99\%$ confidence interval for the estimated probability


\newpage
## Question 2.c

``` {r 2_c_prediction}
# new data to predict
newData = data.frame(
  Sex = rep(c('M', 'F'), c(3, 2)),
  Race = c('white', 'white', 'hispanic', 'black', 'asian'),
  ageC = 0,
  RuralUrban = rep(c('Rural', 'Urban'), c(1, 4))
)

# predicted data
smokePred = as.data.frame(predict(smokeModel, newData,
                                  se.fit = TRUE, type = 'link'))[, 1:2]

smokePred$lower = smokePred$fit - 3 * smokePred$se.fit
smokePred$upper = smokePred$fit + 3 * smokePred$se.fit
smokePred

# predicted odds
expSmokePred = exp(smokePred[, c('fit', 'lower', 'upper')])
knitr::kable(cbind(newData[, -3], 1000 * expSmokePred / (1 + expSmokePred)),
             digits = 1)
```

The claim that rural white males are the group most likely to use
chewing tobacco is likely to be true. This is because the fitted value for rural white male is the highest among all groups. Also, the confidence interval of rural white males do not overlap with that of any other group.

The claim that less than half of one percent of ethnic-minority urban
women and girls chew tobacco is likely to be false given the data. Considering black and asian urban female, the fitted probability of them chewing tobacco is in $99\%$ confidence interval of $(0.13\%, 0.42\%)$ and $(0.08\%, 0.68\%)$. We can see that the $99\%$ confidence interval for black urban women is less than half of one percent. However, the $99\%$ confidence interval for asian urban women includes $0.5\%$ of its population. Thus claim only partially true.


\newpage
# Question 3


## Read Data

``` {r 3_read_data}
# Read the fiji data
fijiFile = 'fijiDownload.RData'
if (!file.exists(fijiFile)) {
  download.file('http://pbrown.ca/teaching/303/data/fiji.RData',
                fijiFile)
}
(load(fijiFile))
glimpse(fiji)
```

``` {r 3_data_sanitization}
# take only children after married and remove na.
fijiSub = fiji[fiji$monthsSinceM > 0 & !is.na(fiji$literacy), ]

# new variables
fijiSub$logYears = log(fijiSub$monthsSinceM / 12)
fijiSub$ageMarried = relevel(fijiSub$ageMarried, '15to18')
fijiSub$urban = relevel(fijiSub$residence, 'rural')
```


\newpage
## Question 3.a

``` {r 3_a_poisson_glm1}
# full poisson glm model
fijiRes = glm(
  children ~ offset(logYears) + ageMarried + ethnicity + literacy + urban,
  family = poisson(link = log),
  data = fijiSub
)
logRateMat = cbind(est = fijiRes$coef, confint(fijiRes, level = 0.99))
knitr::kable(cbind(summary(fijiRes)$coef,
                   exp(logRateMat)),
             digits = 3)
```

The statistical model that corresponds to `fijiRes` is 

$$
Y_{i} \sim Poisson(\lambda)
$$
$$
h(\mu_{i}) = h(\lambda_{i}) = log(\frac{\lambda_{i}}{O_{i}}) = X^{T}_i \beta
$$

where

- $Y_{i}$ is the number of children.
- $\mu_{i}$ is the sample mean $E(Y_i) = E(children)$.
- $O_{i}$ is the offset term, which is the number of years since married.
- $h(\mu_{i})$ is the log link function.

Within $X_{i}$, there is an intercept of all 1's, and indicator variables for Age range of marriage (15-18, 0-15, 18-20, 0-22, 22-25, 25-30, 30+), ethnicity (Fijian, Indian, European, Part European, Pacific Islander, Routman, Chinese, Others) + literacy (Yes, No) + urban (Rural, Suva, Other Urban).

The intercept represents predicted rates of children per month for the subset of females who are between 15 to 18 years old, Fijian, rural and literate.


\newpage
## Question 3.b

``` {r 3_b_poisson_glm2}
# nested poisson glm model
fijiSub$marriedEarly = fijiSub$ageMarried == '0to15'
fijiRes2 = glm(
  children ~ offset(logYears) + marriedEarly + ethnicity + urban,
  family = poisson(link = log),
  data = fijiSub
)
logRateMat2 = cbind(est = fijiRes2$coef, confint(fijiRes2, level = 0.99))
knitr::kable(cbind(summary(fijiRes2)$coef,
                   exp(logRateMat2)),
             digits = 3)
```

The model `fijiRes2` is nested within `fijiRes`.

`fijiRes2` can be viewed as `fijiRes` stripping ethnicity, and only retaining one age group indicator variable. The `marriedEarly` variable can simply be seen as the indicator variable for Age range of marriage 0-15. 

The constraint for the restricted model is that the ages greater than 15 are combined together. All ages above 15 are considered as 1 for the dummy variable.

Thus the coefficients $\beta$ for `fijiRes2` will have 6 less rows and is a subset of the coefficients for `fijiRes`.


\newpage
## Question 3.c
``` {r 3_c_lrtest}
lmtest::lrtest(fijiRes2, fijiRes)
```

From the lmtest result above, we can see that the model being compared is `fijiRes2` and `fijiRes`.

`fijiRes` takes into account for women of different ages and their literacy.

`fijiRes2` is nested within `fijiRes`, and does not account for women of different ages above 15 years old, and their literacy.

The p-value for the test is $0.3834$. This is not significant enough to reject the null hypothesis that adding literacy and age range improves how well the model explains the data.

Thus the first claim is likely to be false, and the second claim is likely to be true.


<!-- \newpage -->
<!-- # Appendix -->

<!-- ```{r ref.label=knitr::all_labels(), echo = T, eval = F} -->
<!-- ``` -->
